Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-21 report/report.tex
Algorithms 1 and 2

\begin{algorithm}
\caption{Regionally Additive Model (RAM) training}
\label{alg:ram}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{A dataset $(X, y)$}
\Output{A trained RAM model $f^{\mathtt{RAM}}$}
  \BlankLine
  Train a differentiable black box model $f$ using $(X, y)$\;
  \tcc{Detect subregions for all features}
  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\}, t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$)\;
  Create the extended feature space $\mathcal{X}^{\mathtt{RAM}}$ using all
  $\mathcal{R}_{st}$, as in Eq. \eqref{eq:ram_feature_space} \;
  Fit a GAM in $\mathcal{X}^{\mathtt{RAM}}$ \tcp*{i.e., train each $f_{st}$ using
    only data in $\mathcal{R}_{st}$}
  \Return{$f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}), \quad \xb \in \Xcal^{\mathtt{RAM}}$}
\end{algorithm}

\begin{algorithm}
\caption{Detection of Subregions using DALE}
\label{alg:subregion_detection}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{Data matrix $X$, Black box model $f$, Maximum Depth $T$}
\BlankLine
\Output{Subregions $\mathcal{R}_{sK}$ for all features $s$ and depth $K=2^T$}
\BlankLine
Initialize empty arrays for \texttt{loss}, \texttt{position}, and \texttt{feature\_c}
\BlankLine
\For{$s = 1$ to $D$}{
X\_list, J\_list = [X], [J] \tcp*{Init lists}
\For{$t = 1$ to $T$}{
    \tcc{Find best split}
    L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
    \tcc{Store best split}
    loss[s, t], position[s, t], feature\_c[s, t] $\gets$ L, p, c
}
}
\Return{loss, position, feature\_c}
\end{algorithm}

>>>>

\begin{algorithm}
\caption{Regionally Additive Model (RAM) training}
\label{alg:ram}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{A dataset $(X, y)$ and a maximum level $T$}
\Output{A trained RAM model $f^{\mathtt{RAM}}$}
  \BlankLine
  Train a differentiable black box model $f$ using $(X, y)$\;
  \tcc{Detect subregions for all features}
  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\},
  t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$, $T$)\;
  Create the extended feature space $\mathcal{X}^{\mathtt{RAM}}$ using all
  $\mathcal{R}_{st}$, as in Eq. \eqref{eq:ram_feature_space} \;
  Fit a GAM in $\mathcal{X}^{\mathtt{RAM}}$ \tcp*{i.e., train each $f_{st}$ using
    only data in $\mathcal{R}_{st}$}
  \Return{$f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}), \quad \xb \in \Xcal^{\mathtt{RAM}}$}
\end{algorithm}

\begin{algorithm}
\caption{Detection of Subregions using DALE}
\label{alg:subregion_detection}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{Dataset $(X, y)$, Black box model $f$, Maximum depth $T$}
\BlankLine
\Output{Subregions $\mathcal{R}_{sT_s}$ for all features $s$, where $T_s \leq 2^T$}
\BlankLine
\tcc{Needed for DALE feature effect estimation}
Compute the Jacobian w.r.t. features $\xb$,  $J = \nabla_{\xb}f(\xb)$ \;
\For{$s = 1$ to $D$}{
  \For{$t = 1$ to $T$}{
    \tcc{Find best split feature $c$ at point $p$, leading to loss
      $\mathcal{H}_{st}$ using regions of previous level}
    $\mathcal{H}_{st}, c, p$ $\gets$
    BestSplit($\left\{\mathcal{R}_{sk}|k\in\left\{1,\ldots,T_{t-1}\right\}\right\}$)\;
    \If{$\lVert\mathcal{H}_{st} - \mathcal{H}_{s(t-1)}\rVert < \epsilon$}{
      break\;
    }
  }
}
\Return{$\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\},
  t\in\left\{1,\ldots,T_s\right\}\right\}$}
\end{algorithm}
